---
title: "Lab 7 Spring 2025"
author: "Stephanie Gonzalez Gil"
date: "2025-03-20"
output: word_document
editor_options: 
  markdown: 
    wrap: 72
---

# Agenda

This lab session has three components:

1.  A basic view of GitHub

2.  Key data management and cleaning codes

3.  Loop tutorial

# Section 1: GitHub

We'll start by looking at GitHub.

# Section 2: Data Management

Next, we are going to move on to some basic data cleaning commands that will help you as you start to collect data.

**NOTE**: Never make changes in your Excel file. Only make changes using
your R Code.

## Missing Data

Let's begin by dealing with missing data. Missing data can cause
problems like:

-   Reduction in sample size due to unobserved values
-   Biased results due to systematic differences between observed and
    missing data
-   No replacement for observed data

It is important to consider the amount, measure, pattern, and possible
causes of missing data when we are trying to manage it.

There are three main types:

1.  Missing completely at random (MCAR), no systemic missing data. *This
    is very unlikely*
2.  Missing at random (MAR), systemic missing data of observed and
    unobserved data. *We are able to account for this in analysis*.
3.  Missing not at random (MNAR), systemic missing data of unobserved
    data. *We cannot account for this in analysis.*

There are two main measures of "missingness":

1.  Percentage missing per variable
2.  Pairwise missing rate

There are multiple ways to manage missing data:

-   Complete case deletion

-   Simple substitution (replace value with summary statistic)

-   Assigning separate category

-   Multiple imputation (missing values are replaced by a random sample
    of plausible values)

R Studio also offers us tools to visualize missing patterns with the
mice package using the codes md.pattern(), and md.pair().

-   mice package includes:

    -   md.pattern()

    -   md.pair()

```{r}
#install.packages("mice")
library(mice)
View(airquality)
?airquality 
```

Let's use md.pattern() first

```{r}
?md.pattern
md.pattern(airquality)
```

Now let's look at md.pairs()

```{r}
?md.pairs
md.pairs(airquality)
```

Now that we know we have missing data how do we manage it? One of the
things we can do is omit the missing variables altogether.

```{r}
ds <- airquality
ds_noNA <- na.omit(airquality)
dim(ds) - dim(ds_noNA)
#We have removed 42 observations. 
```

With missing data removed from our data, we are now ready to create our
working data set and rename our variables if necessary. These commands
should be review.

```{r}
library(dplyr)
ds_working <- ds_noNA %>% 
  select(Ozone,Solar.R,Temp, Month)

ds_working_2 <- ds_working %>%
  rename(ozone = Ozone) %>% 
  rename(sol_rad = Solar.R) %>% 
  rename(temp = Temp) %>% 
  rename(month = Month)

View(ds_working_2)
```

Now let's say I am interested in making Temp a categorical variable
using the quantiles of the data.

```{r}
quant <- quantile(ds_working_2$temp)
quant

ds_final<- ds_working_2 %>%
  mutate(temp.cat = findInterval(temp, quantile(temp), 
                                    rightmost.closed = TRUE))

ds_final$temp_cat <- as.factor(ds_final$temp.cat)

View(ds_final)
```

Now that I have the missing values removed, only the variables I want
named the way I want, and my new categorical variable for analysis,
let's see how I can use R to help me make my Table 1.

```{r}
#install.packages("tableone")
library(tableone)
x <- CreateTableOne(data = ds_final,
                    vars = c("ozone","sol_rad","temp_cat"))
x
```
             
# Section 3: Loop Functions

Loop is a base r command that runs commands for each elements
(observation or value) of a vector, and returns a string of results.

The loop function, has a basic syntax:

for (<element> in <a string of elements>) {
<vector to contain our loop outputs>[element] \<-
function(<for each element>) }

It will take one <element> from our <string of elements>, run the
function specified in the function() statement, return the output into
the <vector to contain our loop outputs>, then repeat the step for the
following <element> in the <string of elements> until all are computed.

<element> can be any name but usually use 'i' <string of elements> can
be a vector that points to the order of elements to be processed.
function() can be multiple written in terms of the computation for each
element

Let's break down the components:

1\. for (<element> in <a string of elements>)

{ 2. function(<for each element>)

3\. <vector to contain our loop outputs>[element] }

4\. initiation of <vector to contain our loop outputs> before running
the loop to store the results

**Example:** We will create a dataframe 'musical_db', and run a simple
loop on each of the values for one of the variables 'noise_level' to
create a new variable 'power' that will contain the results of (each
value of noise_level)\*100

```{r}
noise_level <- c(20, 30, 50)
instruments <- c("violin", "piano", "cello")

musical_db <- data.frame(instruments, noise_level)

View(musical_db)

```

So, let's disassemble the steps,

first, lets state our

for (<element> in <a string of elements>)

we are running the computation for each <element> or observation of
noise level,

we will code this by using the length statement for our example (note we
can also use levels in factor variables with levels, or other metrics)

```{r}

length(musical_db$noise_level)

```

we find we have 3 observations of noise_level we will be computing in
turn.

We plan to run a loop that is read as "for all elements i, from 1 to
length of the specified variable number of observations..."

we write this as:

```         
for (i in 1:length(musical_db$noise_level))
```

Next, we define the function

we want to compute for each observation of noise level, noise_level\*100

so our function is:

noise_level[i]\*100

so far, we have defined our <element>, <string of elements>, the
function()

now lets assign a variable to store the outputs:

music_db\$power \<- noise_level[i]\*100

Now, let's assemble these components:

```{r}
#initiating the dummy variable to store our output
musical_db$power<-c()


for (i in 1:length(musical_db$noise_level))
  {
musical_db$power[i]<- noise_level[i]*100
}

#now lets view our resulting dataframe
View(musical_db)
```

On a side note, we can use the function command to create a customized
command to carry out

For example, let's create a function 'x' that multiplies each value by
10\^3, we can run the function on all values of noise_level to create a
new variable "powered_3"

```{r}
x <- function(i) i*10^3

musical_db$powered3 <- x(musical_db$noise_level)

View(musical_db)
```

Let's look at our malawi dataset to see how this could be useful for
Problem Set #2. We are going to examine radio ownership (hv207) by
wealth index (hv270) as well as tv ownership (hv208) by wealth index.

```{r}
malawi <- read.csv("~/Documents/GH 811/PS2/malawi2010.csv")

View(malawi)

ds_working <- malawi %>% select(hv207,hv208,hv270)

ds_noNA <- ds_working %>% 
  filter(hv207!=9) %>% 
  filter(hv208!=9) %>% 
  filter(hv270!=9)

ds_final <- ds_noNA %>% 
  rename(radio = hv207) %>% 
  rename(wealth = hv270) %>% 
  rename(tv = hv208)

ds_final$wealth.cat <- ordered(ds_final$wealth, levels = c("poorest","poorer","middle","richer","richest"))

ds_final$radio.f <- ordered(ds_final$radio, levels = c(1,0))
head(ds_final)
```

Now, let's start building our loop.

```{r}
#creating vector 'x' to contain (concatenate) our loop output
x <- c()

#viewing our dummy vector
x

#checking our levels 
levels(ds_final$wealth.cat)
```

Let's recall,

The loop function, has a specific syntax:

for (<element> in <a string of elements>) {
<vector to contain our loop outputs>[element] \<-
function(<for each element>) }

```{r}
for (i in levels(ds_final$wealth.cat))
  {
  x[i] <-
   prop.table(table(ds_final$radio.f[ds_final$wealth.cat==i]))[1]*100
  }

x

```

We can check our work with prop.table

```{r}
y <- prop.table(table(ds_final$wealth.cat, ds_final$radio), 1)*100
y
```

Now, lets generate a bar chart using the results.

```{r}
barplot(x,
main = "Proportion of Radio Use by Wealth Status, Malawi",
xlab="Wealth Index", 
ylab="Proportion of Radio Use(%)",
ylim=c(0,100))
```

Now create a for loop and bar chart for TV use. What do you see?

```{r}
ds_final$tv.f <- ordered(ds_final$tv, levels=c(1,0))
head(ds_final)

for (i in levels(ds_final$wealth.cat)){
  x[i] <-
   prop.table(table(ds_final$tv.f[ds_final$wealth.cat==i]))[1]*100
 }
x

y<- prop.table(table(ds_final$wealth.cat,ds_final$tv.f),1)*100
y

barplot(x,
main="Proportion of TV Use by Wealth Status, Malawi",
xlab="Wealth Index", 
ylab="Proportion of TV Use(%)",
ylim=c(0, 100))
```
